---
title: "lps7.1"
output: html_document
date: "2026-03-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(Matrix)
library(readr)
library(data.table)
library(fastTopics)
library(ebnm)
library(flashier)
library(ggplot2)
library(ggrepel)
library(cowplot)
library(rsvd)
library(uwot)

library(limma)
library(matrixStats)
library(ashr)
library(dplyr)
library(tidyr)
library(reshape2)
```

## 1. Read and preprocess LPS data
```{r}
read_lps_data <- function (file) {
  counts <- fread(file)
  class(counts) <- "data.frame"
  genes <- counts[,1]
  counts <- t(as.matrix(counts[,-1]))
  colnames(counts) <- genes
  samples <- rownames(counts)
  samples <- strsplit(samples,"_")
  samples <- data.frame(tissue    = sapply(samples,"[[",1),
                        timepoint = sapply(samples,"[[",2),
                        mouse     = sapply(samples,"[[",3))
  samples <- transform(samples,
                       tissue    = factor(tissue),
					   timepoint = factor(timepoint),
					   mouse     = factor(mouse))
  return(list(samples = samples,counts = counts))
}
out <- read_lps_data("../data/lps.csv.gz")
samples <- out$samples
counts  <- out$counts
rm(out)

dim(samples)
dim(counts)
#column = one gene,
counts
samples
```

Remove a sample that appears to be an outlier based on the NMF analyses:

```{r filter-outliers}
i       <- which(rownames(counts) != "iLN_d2_20")
samples <- samples[i,]
counts  <- counts[i,]
```

Remove genes that are expressed in fewer than 5 samples:

```{r filter-genes}
j <- which(colSums(counts > 0) > 4)
counts <- counts[,j]
```


## 2. Fit topic model and structure plot


```{r fit-topic-model, cache=TRUE}
set.seed(1)
tm <- fit_poisson_nmf(counts,k = 14,init.method = "random",method = "em",
                      numiter = 20,verbose = "none",
					  control = list(numiter = 4,nc = 8,extrapolate = FALSE))
tm <- fit_poisson_nmf(counts,fit0 = tm,method = "scd",numiter = 40,
                      control = list(numiter = 4,nc = 8,extrapolate = TRUE),
					  verbose = "none")
tm_mn <- poisson2multinom(tm)
L <- tm_mn$L 
```


## 3. Build pairwise log-ratios and fit limma

```{r}
tp <- factor(
  as.character(samples$timepoint),
  levels = c("d0","d0.25","d0.5","d1","d2","d3","d5")
)
tp <- relevel(tp, ref = "d0")

keep <- !is.na(tp)
samples_sub <- droplevels(samples[keep, , drop = FALSE])
L_sub <- L[keep, , drop = FALSE]
tp <- droplevels(tp[keep])

N <- nrow(L_sub)
K <- ncol(L_sub)

# topic-specific epsilon
c0 <- 1e-4
med_k <- colMedians(L_sub, na.rm = TRUE)
eps_k <- c0 * med_k
eps_k <- pmin(pmax(eps_k, 1e-12), 1e-2)

# clamp to [0,1]
L01 <- pmin(pmax(L_sub, 0), 1)

pair_mat <- t(combn(K, 2))
n_pairs <- nrow(pair_mat)

log_ratio_mat <- matrix(NA_real_, nrow = N, ncol = n_pairs)
for (p in seq_len(n_pairs)) {
  k1 <- pair_mat[p, 1]
  k2 <- pair_mat[p, 2]
  log_ratio_mat[, p] <- log(L01[, k1] + eps_k[k1]) -
                        log(L01[, k2] + eps_k[k2])
}

colnames(log_ratio_mat) <- apply(pair_mat, 1, function(idx) {
  paste0("T", idx[1], "_vs_T", idx[2])
})

Y_lr <- t(log_ratio_mat)

design <- model.matrix(~ tp + tissue, data = samples_sub)
colnames(design) <- make.names(colnames(design))

fit <- lmFit(Y_lr, design)
fit <- eBayes(fit)
```


## 4. Pairwise significance heatmap input

```{r}
coef_names <- colnames(fit$p.value)
tp_coefs <- grep("^tp", coef_names, value = TRUE)

pair_names <- rownames(fit$p.value)
parts <- do.call(rbind, strsplit(pair_names, "_vs_"))
k1 <- as.integer(sub("^T", "", parts[, 1]))
k2 <- as.integer(sub("^T", "", parts[, 2]))

K <- max(c(k1, k2))

base_grid <- expand.grid(k1 = 1:K, k2 = 1:K)
base_grid <- base_grid[base_grid$k1 != base_grid$k2, , drop = FALSE]

make_grid_for_coef <- function(coef_name) {
  pvals <- fit$p.value[, coef_name]
  qvals <- p.adjust(pvals, method = "BH")
  sig <- qvals < 0.05

  df_pairs <- data.frame(k1 = k1, k2 = k2, sig = sig)

  m1 <- merge(base_grid, df_pairs, by = c("k1", "k2"), all.x = TRUE)
  df_swapped <- transform(df_pairs, k1 = k2, k2 = k1)
  m2 <- merge(base_grid, df_swapped, by = c("k1", "k2"), all.x = TRUE)

  out <- base_grid
  out$sig <- ifelse(!is.na(m1$sig), m1$sig, m2$sig)
  out$coef <- coef_name
  out
}

grid_all <- do.call(rbind, lapply(tp_coefs, make_grid_for_coef))
```



## 5. Compare ordinary, limma, and ASH

```{r}
fit0 <- lmFit(Y_lr, design)

coef_hat <- fit0$coefficients
se_hat <- sweep(fit0$stdev.unscaled, 1, fit0$sigma, "*")

t_ord <- coef_hat / se_hat
p_ord <- 2 * pt(-abs(t_ord), df = fit0$df.residual)

fit_eb <- eBayes(fit0)

df_resid <- unique(fit0$df.residual)

ash_list <- lapply(tp_coefs, function(coef_name) {
  j <- match(coef_name, colnames(coef_hat))
  ash(coef_hat[, j], se_hat[, j], df = df_resid)
})
names(ash_list) <- tp_coefs
```


```{r}
q_ord <- apply(p_ord[, tp_coefs, drop = FALSE], 2, p.adjust, method = "BH")
q_limma <- apply(fit_eb$p.value[, tp_coefs, drop = FALSE], 2, p.adjust, method = "BH")

q_ash_mat <- sapply(ash_list, function(a) get_qvalue(a))
sig_ash <- q_ash_mat < 0.05
colnames(sig_ash) <- tp_coefs
rownames(sig_ash) <- rownames(Y_lr)
```


```{r}
state_mat <- sapply(tp_coefs, function(coef_name) {
  paste0(
    ifelse(q_ord[, coef_name]   < 0.05, "N", "-"),
    ifelse(q_limma[, coef_name] < 0.05, "L", "-"),
    ifelse(sig_ash[, coef_name],        "A", "-")
  )
})

colnames(state_mat) <- tp_coefs
rownames(state_mat) <- rownames(Y_lr)

state_levels <- c("---", "N--", "-L-", "--A", "NL-", "N-A", "-LA", "NLA")

state_frac <- sapply(tp_coefs, function(coef_name) {
  tab <- table(factor(state_mat[, coef_name], levels = state_levels))
  as.numeric(tab) / nrow(state_mat)
})

colnames(state_frac) <- tp_coefs
rownames(state_frac) <- state_levels
```


```{r}
df_plot <- reshape2::melt(state_frac)
colnames(df_plot) <- c("state", "coefficient", "fraction")

ggplot(df_plot, aes(x = state, y = coefficient, fill = fraction)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(
    title = "Agreement of pairwise significance calls across three methods",
    x = "Significance state (N=ordinary, L=limma, A=ASH)",
    y = "Coefficient",
    fill = "Fraction of pairs"
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


## 6. Reusable anchor-topic trajectory helpers

```{r}
time_map <- c(
  "tpd0.25" = 0.25,
  "tpd0.5"  = 0.5,
  "tpd1"    = 1,
  "tpd2"    = 2,
  "tpd3"    = 3,
  "tpd5"    = 5
)

make_anchor_pair_info <- function(anchor_topic, Y_lr) {
  pair_keep <- grepl(
    paste0("^T", anchor_topic, "_vs_T|_vs_T", anchor_topic, "$"),
    rownames(Y_lr)
  )
  pair_names_anchor <- rownames(Y_lr)[pair_keep]

  parts_anchor <- do.call(rbind, strsplit(pair_names_anchor, "_vs_"))
  left_topic  <- as.integer(sub("^T", "", parts_anchor[, 1]))
  right_topic <- as.integer(sub("^T", "", parts_anchor[, 2]))

  flip_sign <- (right_topic == anchor_topic)
  other_topic <- ifelse(left_topic == anchor_topic, right_topic, left_topic)

  pair_info <- data.frame(
    pair = pair_names_anchor,
    left_topic = left_topic,
    right_topic = right_topic,
    other_topic = other_topic,
    flip_sign = flip_sign,
    pair_label = paste0("T", anchor_topic, "_vs_T", other_topic),
    stringsAsFactors = FALSE
  )

  list(pair_keep = pair_keep, pair_names_anchor = pair_names_anchor, pair_info = pair_info)
}

make_limma_anchor_df <- function(anchor_topic, Y_lr, fit_eb, tp_coefs, time_map) {
  anchor_obj <- make_anchor_pair_info(anchor_topic, Y_lr)
  pair_keep <- anchor_obj$pair_keep
  pair_names_anchor <- anchor_obj$pair_names_anchor
  pair_info <- anchor_obj$pair_info

  coef_limma_mat <- fit_eb$coefficients[pair_keep, tp_coefs, drop = FALSE]

  q_limma2 <- sapply(tp_coefs, function(coef_name) {
    p.adjust(fit_eb$p.value[pair_keep, coef_name], method = "BH")
  })
  colnames(q_limma2) <- tp_coefs
  rownames(q_limma2) <- pair_names_anchor

  df_limma <- expand.grid(
    pair = pair_names_anchor,
    coef_name = tp_coefs,
    stringsAsFactors = FALSE
  )

  df_limma <- merge(df_limma, pair_info, by = "pair", all.x = TRUE)

  df_limma$estimate_raw <- mapply(function(p, c) coef_limma_mat[p, c],
                                  df_limma$pair, df_limma$coef_name)

  df_limma$estimate <- ifelse(df_limma$flip_sign,
                              -df_limma$estimate_raw,
                              df_limma$estimate_raw)

  df_limma$signif <- mapply(function(p, c) q_limma2[p, c] < 0.05,
                            df_limma$pair, df_limma$coef_name)

  df_limma$time <- unname(time_map[df_limma$coef_name])
  df_limma$method <- "limma"
  df_limma
}

make_ash_anchor_df <- function(anchor_topic, Y_lr, coef_hat, ash_list, tp_coefs, time_map, use_pm = FALSE) {
  anchor_obj <- make_anchor_pair_info(anchor_topic, Y_lr)
  pair_keep <- anchor_obj$pair_keep
  pair_names_anchor <- anchor_obj$pair_names_anchor
  pair_info <- anchor_obj$pair_info

  names(ash_list) <- tp_coefs

  q_ash_mat <- do.call(cbind, lapply(tp_coefs, function(coef_name) {
    get_qvalue(ash_list[[coef_name]])[pair_keep]
  }))
  colnames(q_ash_mat) <- tp_coefs
  rownames(q_ash_mat) <- pair_names_anchor

  if (use_pm) {
    coef_ash_mat <- do.call(cbind, lapply(tp_coefs, function(coef_name) {
      get_pm(ash_list[[coef_name]])[pair_keep]
    }))
    colnames(coef_ash_mat) <- tp_coefs
    rownames(coef_ash_mat) <- pair_names_anchor
  } else {
    coef_ash_mat <- coef_hat[pair_keep, tp_coefs, drop = FALSE]
  }

  df_ash <- expand.grid(
    pair = pair_names_anchor,
    coef_name = tp_coefs,
    stringsAsFactors = FALSE
  )

  df_ash <- merge(df_ash, pair_info, by = "pair", all.x = TRUE)

  df_ash$estimate_raw <- mapply(function(p, c) coef_ash_mat[p, c],
                                df_ash$pair, df_ash$coef_name)

  df_ash$estimate <- ifelse(df_ash$flip_sign,
                            -df_ash$estimate_raw,
                            df_ash$estimate_raw)

  df_ash$signif <- mapply(function(p, c) q_ash_mat[p, c] < 0.05,
                          df_ash$pair, df_ash$coef_name)

  df_ash$time <- unname(time_map[df_ash$coef_name])
  df_ash$method <- "ASH"
  df_ash
}

plot_anchor_df <- function(df_in, anchor_topic, method_label) {
  ggplot(df_in, aes(x = time, y = estimate, group = pair_label)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
    geom_line(color = "black") +
    geom_point(aes(fill = signif), shape = 21, color = "black", size = 1.5) +
    facet_wrap(~ pair_label, scales = "free_y", ncol = 4) +
    scale_fill_manual(values = c("TRUE" = "black", "FALSE" = "white"),
                      name = ifelse(method_label == "limma", "BH < 0.05", "ASH q < 0.05")) +
    labs(
      title = paste0("Anchor topic T", anchor_topic, " vs all others (", method_label, ")"),
      x = "Time",
      y = paste0("Estimated effect on log(T", anchor_topic, "/other) vs d0")
    ) +
    theme_bw(base_size = 13)
}
```


## 7. Anchor topic plots for T1 and T7

```{r}
anchor_topic <- 1
df_limma_T1 <- make_limma_anchor_df(anchor_topic, Y_lr, fit_eb, tp_coefs, time_map)
df_ash_T1   <- make_ash_anchor_df(anchor_topic, Y_lr, coef_hat, ash_list, tp_coefs, time_map, use_pm = FALSE)

plot_anchor_df(df_limma_T1, anchor_topic, "limma")
plot_anchor_df(df_ash_T1, anchor_topic, "ASH")
```


```{r}
anchor_topic <- 7
df_limma_T7 <- make_limma_anchor_df(anchor_topic, Y_lr, fit_eb, tp_coefs, time_map)
df_ash_T7   <- make_ash_anchor_df(anchor_topic, Y_lr, coef_hat, ash_list, tp_coefs, time_map, use_pm = FALSE)

plot_anchor_df(df_limma_T7, anchor_topic, "limma")
plot_anchor_df(df_ash_T7, anchor_topic, "ASH")
```