---
title: "lps7.3"
output: pdf_document
date: "2026-04-09"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(Matrix)
library(readr)
library(data.table)
library(fastTopics)
library(ebnm)
library(flashier)
library(ggplot2)
library(ggrepel)
library(cowplot)
library(rsvd)
library(uwot)

library(limma)
library(matrixStats)
library(ashr)
library(dplyr)
library(tidyr)
library(reshape2)
library(mr.ashr)
```



## 1. Read and preprocess LPS data
```{r}
read_lps_data <- function (file) {
  counts <- fread(file)
  class(counts) <- "data.frame"
  genes <- counts[,1]
  counts <- t(as.matrix(counts[,-1]))
  colnames(counts) <- genes
  samples <- rownames(counts)
  samples <- strsplit(samples,"_")
  samples <- data.frame(tissue    = sapply(samples,"[[",1),
                        timepoint = sapply(samples,"[[",2),
                        mouse     = sapply(samples,"[[",3))
  samples <- transform(samples,
                       tissue    = factor(tissue),
					   timepoint = factor(timepoint),
					   mouse     = factor(mouse))
  return(list(samples = samples,counts = counts))
}
out <- read_lps_data("../data/lps.csv.gz")
samples <- out$samples
counts  <- out$counts
rm(out)

dim(samples)
dim(counts)
#column = one gene,
counts
samples
```

Remove a sample that appears to be an outlier based on the NMF analyses:

```{r filter-outliers}
i       <- which(rownames(counts) != "iLN_d2_20")
samples <- samples[i,]
counts  <- counts[i,]
```

Remove genes that are expressed in fewer than 5 samples:

```{r filter-genes}
j <- which(colSums(counts > 0) > 4)
counts <- counts[,j]
```


## 2. Fit topic model and structure plot


```{r fit-topic-model, cache=TRUE}
set.seed(1)
tm <- fit_poisson_nmf(counts,k = 14,init.method = "random",method = "em",
                      numiter = 20,verbose = "none",
					  control = list(numiter = 4,nc = 8,extrapolate = FALSE))
tm <- fit_poisson_nmf(counts,fit0 = tm,method = "scd",numiter = 40,
                      control = list(numiter = 4,nc = 8,extrapolate = TRUE),
					  verbose = "none")
tm_mn <- poisson2multinom(tm)
L <- tm_mn$L 
```


## 3. Build pairwise log-ratios and fit limma

```{r}
tp <- factor(
  as.character(samples$timepoint),
  levels = c("d0","d0.25","d0.5","d1","d2","d3","d5")
)
tp <- relevel(tp, ref = "d0")

keep <- !is.na(tp)
samples_sub <- droplevels(samples[keep, , drop = FALSE])
L_sub <- L[keep, , drop = FALSE]
tp <- droplevels(tp[keep])

N <- nrow(L_sub)
K <- ncol(L_sub)

# topic-specific epsilon
c0 <- 1e-4
med_k <- colMedians(L_sub, na.rm = TRUE)
eps_k <- c0 * med_k
eps_k <- pmin(pmax(eps_k, 1e-12), 1e-2)

# clamp to [0,1]
L01 <- pmin(pmax(L_sub, 0), 1)

pair_mat <- t(combn(K, 2))
n_pairs <- nrow(pair_mat)

log_ratio_mat <- matrix(NA_real_, nrow = N, ncol = n_pairs)
for (p in seq_len(n_pairs)) {
  k1 <- pair_mat[p, 1]
  k2 <- pair_mat[p, 2]
  log_ratio_mat[, p] <- log(L01[, k1] + eps_k[k1]) -
                        log(L01[, k2] + eps_k[k2])
}

colnames(log_ratio_mat) <- apply(pair_mat, 1, function(idx) {
  paste0("T", idx[1], "_vs_T", idx[2])
})

Y_lr <- t(log_ratio_mat)

design <- model.matrix(~ tp + tissue, data = samples_sub)
colnames(design) <- make.names(colnames(design))

fit <- lmFit(Y_lr, design)
fit <- eBayes(fit)
```


## 4. Pairwise significance heatmap input

```{r}
coef_names <- colnames(fit$p.value)
tp_coefs <- grep("^tp", coef_names, value = TRUE)

pair_names <- rownames(fit$p.value)
parts <- do.call(rbind, strsplit(pair_names, "_vs_"))
k1 <- as.integer(sub("^T", "", parts[, 1]))
k2 <- as.integer(sub("^T", "", parts[, 2]))

K <- max(c(k1, k2))

base_grid <- expand.grid(k1 = 1:K, k2 = 1:K)
base_grid <- base_grid[base_grid$k1 != base_grid$k2, , drop = FALSE]

make_grid_for_coef <- function(coef_name) {
  pvals <- fit$p.value[, coef_name]
  qvals <- p.adjust(pvals, method = "BH")
  sig <- qvals < 0.05

  df_pairs <- data.frame(k1 = k1, k2 = k2, sig = sig)

  m1 <- merge(base_grid, df_pairs, by = c("k1", "k2"), all.x = TRUE)
  df_swapped <- transform(df_pairs, k1 = k2, k2 = k1)
  m2 <- merge(base_grid, df_swapped, by = c("k1", "k2"), all.x = TRUE)

  out <- base_grid
  out$sig <- ifelse(!is.na(m1$sig), m1$sig, m2$sig)
  out$coef <- coef_name
  out
}

grid_all <- do.call(rbind, lapply(tp_coefs, make_grid_for_coef))
```

## 5. Compare ordinary, limma, ASH, EPB, and MR.ASHR
```{r}

library(JuliaCall)
library(mr.ashr)
library(reshape2)
library(ggplot2)

julia_setup(JULIA_HOME = "/opt/anaconda3/julia_env/pyjuliapkg/install/bin")
julia_command('import Pkg')
julia_command('Pkg.activate("/opt/anaconda3/julia_env")')
julia_command("using Empirikos, Hypatia, MultipleTesting")

run_epb_one_coef <- function(coef_name, coef_hat, se_hat, df_resid, alpha = 0.05) {
  beta_vec <- as.numeric(coef_hat[, coef_name])
  se2_vec  <- as.numeric(se_hat[, coef_name]^2)
  nu_val   <- as.integer(df_resid[1])

  julia_assign("beta_vec_r", beta_vec)
  julia_assign("se2_vec_r", se2_vec)
  julia_assign("nu_val_r", nu_val)
  julia_assign("alpha_r", alpha)

  julia_command("samples_epb = [Empirikos.NormalChiSquareSample(beta_vec_r[i], se2_vec_r[i], nu_val_r) for i in eachindex(beta_vec_r)]")
  julia_command("test_epb_r = Empirikos.EmpiricalPartiallyBayesTTest(α = alpha_r, solver = Hypatia.Optimizer)")
  julia_command("res_epb_r = fit(test_epb_r, samples_epb)")

  list(
    adjp   = as.numeric(julia_eval("res_epb_r.adjp")),
    rj_idx = as.logical(julia_eval("res_epb_r.rj_idx"))
  )
}

fit0 <- lmFit(Y_lr, design)

coef_hat <- fit0$coefficients
se_hat   <- sweep(fit0$stdev.unscaled, 1, fit0$sigma, "*")
df_resid <- unique(fit0$df.residual)

## ordinary
t_ord <- coef_hat / se_hat
p_ord <- 2 * pt(-abs(t_ord), df = fit0$df.residual)
q_ord <- apply(p_ord[, tp_coefs, drop = FALSE], 2, p.adjust, method = "BH")

## limma
fit_eb <- eBayes(fit0)
q_limma <- apply(fit_eb$p.value[, tp_coefs, drop = FALSE], 2, p.adjust, method = "BH")
sig_limma <- q_limma < 0.05

## ASH
ash_list <- lapply(tp_coefs, function(coef_name) {
  j <- match(coef_name, colnames(coef_hat))
  ash(coef_hat[, j], se_hat[, j], df = df_resid)
})
names(ash_list) <- tp_coefs

q_ash_mat <- sapply(ash_list, get_qvalue)
sig_ash <- q_ash_mat < 0.05
colnames(sig_ash) <- tp_coefs
rownames(sig_ash) <- rownames(Y_lr)

## EPB
epb_list <- lapply(tp_coefs, function(coef_name) {
  message("Running EPB for ", coef_name)
  run_epb_one_coef(coef_name, coef_hat, se_hat, df_resid, alpha = 0.05)
})
names(epb_list) <- tp_coefs

q_epb_mat <- sapply(tp_coefs, function(coef_name) epb_list[[coef_name]]$adjp)
sig_epb   <- sapply(tp_coefs, function(coef_name) epb_list[[coef_name]]$rj_idx)

colnames(q_epb_mat) <- tp_coefs
colnames(sig_epb)   <- tp_coefs
rownames(q_epb_mat) <- rownames(Y_lr)
rownames(sig_epb)   <- rownames(Y_lr)

## MR.ASHR
# X_mr <- design[, colnames(design) != "(Intercept)", drop = FALSE]
X_full <- model.matrix(~ tp + tissue, data = samples_sub)
X_mr <- X_full[, colnames(X_full) != "(Intercept)", drop = FALSE]

run_mr_ash_one_pair <- function(y, X) {
  y <- as.numeric(y)

  fit0_mr <- init_mr_ash(
    X = X,
    y = y,
    init.method = "null"
  )

  fit_mr_ash(
    X = X,
    y = y,
    fit0 = fit0_mr,
    standardize = TRUE,
    intercept = TRUE,
    verbose = "none"
  )
}

mr_fits <- lapply(seq_len(nrow(Y_lr)), function(i) {
  message("Running MR.ASHR for ", rownames(Y_lr)[i])
  run_mr_ash_one_pair(Y_lr[i, ], X_mr)
})
names(mr_fits) <- rownames(Y_lr)

mr_coef_mat <- do.call(rbind, lapply(mr_fits, function(fit) {
  c("(Intercept)" = fit$intercept, fit$b)
}))
rownames(mr_coef_mat) <- rownames(Y_lr)
colnames(mr_coef_mat) <- c("(Intercept)", colnames(X_mr))

mr_lfsr_mat <- do.call(rbind, lapply(mr_fits, function(fit) {
  c("(Intercept)" = NA_real_, fit$lfsr)
}))
rownames(mr_lfsr_mat) <- rownames(Y_lr)
colnames(mr_lfsr_mat) <- c("(Intercept)", colnames(X_mr))

mr_tp_coefs <- grep("^tp", colnames(mr_lfsr_mat), value = TRUE)
sig_mr <- mr_lfsr_mat[, mr_tp_coefs, drop = FALSE] < 0.05
```

## five-method agreement heatmap
```{r}
common_tp_coefs <- Reduce(intersect, list(
  colnames(q_ord),
  colnames(q_limma),
  colnames(sig_ash),
  colnames(sig_epb),
  colnames(sig_mr)
))

q_ord_use   <- q_ord[, common_tp_coefs, drop = FALSE]
q_limma_use <- q_limma[, common_tp_coefs, drop = FALSE]
sig_ash_use <- sig_ash[, common_tp_coefs, drop = FALSE]
sig_epb_use <- sig_epb[, common_tp_coefs, drop = FALSE]
sig_mr_use  <- sig_mr[, common_tp_coefs, drop = FALSE]

state_mat5 <- sapply(common_tp_coefs, function(coef_name) {
  paste0(
    ifelse(q_ord_use[, coef_name]   < 0.05, "N", "-"),
    ifelse(q_limma_use[, coef_name] < 0.05, "L", "-"),
    ifelse(sig_ash_use[, coef_name],        "A", "-"),
    ifelse(sig_epb_use[, coef_name],        "E", "-"),
    ifelse(sig_mr_use[, coef_name],         "M", "-")
  )
})

colnames(state_mat5) <- common_tp_coefs
rownames(state_mat5) <- rownames(Y_lr)

all_states5 <- apply(expand.grid(
  N = c("-", "N"),
  L = c("-", "L"),
  A = c("-", "A"),
  E = c("-", "E"),
  M = c("-", "M")
), 1, paste0, collapse = "")

state_frac5 <- sapply(common_tp_coefs, function(coef_name) {
  tab <- table(factor(state_mat5[, coef_name], levels = all_states5))
  as.numeric(tab) / nrow(state_mat5)
})

colnames(state_frac5) <- common_tp_coefs
rownames(state_frac5) <- all_states5

df_plot5 <- melt(state_frac5)
colnames(df_plot5) <- c("state", "coefficient", "fraction")
df_plot5$fraction_fill <- ifelse(df_plot5$fraction == 0, NA, df_plot5$fraction)

ggplot(df_plot5, aes(x = state, y = coefficient, fill = fraction_fill)) +
  geom_tile(color = "white") +
  scale_fill_gradient(
    low = "#d9ecff",
    high = "steelblue",
    na.value = "grey90",
    name = "Fraction of pairs"
  ) +
  labs(
    title = "Agreement across five methods",
    x = "State (N=ordinary, L=limma, A=ASH, E=EPB, M=MR.ASHR)",
    y = "Coefficient"
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))
```



## Reusable anchor-topic trajectory helpers
```{r}
time_map <- c(
  "tpd0.25" = 0.25,
  "tpd0.5"  = 0.5,
  "tpd1"    = 1,
  "tpd2"    = 2,
  "tpd3"    = 3,
  "tpd5"    = 5
)

make_anchor_pair_info <- function(anchor_topic, Y_lr) {
  pair_keep <- grepl(
    paste0("^T", anchor_topic, "_vs_T|_vs_T", anchor_topic, "$"),
    rownames(Y_lr)
  )
  pair_names_anchor <- rownames(Y_lr)[pair_keep]

  parts_anchor <- do.call(rbind, strsplit(pair_names_anchor, "_vs_"))
  left_topic  <- as.integer(sub("^T", "", parts_anchor[, 1]))
  right_topic <- as.integer(sub("^T", "", parts_anchor[, 2]))

  pair_info <- data.frame(
    pair = pair_names_anchor,
    other_topic = ifelse(left_topic == anchor_topic, right_topic, left_topic),
    flip_sign = (right_topic == anchor_topic),
    stringsAsFactors = FALSE
  )

  pair_info$pair_label <- paste0("T", anchor_topic, "_vs_T", pair_info$other_topic)

  list(pair_keep = pair_keep, pair_names_anchor = pair_names_anchor, pair_info = pair_info)
}

make_anchor_df <- function(anchor_topic, Y_lr, coef_mat, signif_mat, tp_coefs, time_map) {
  anchor_obj <- make_anchor_pair_info(anchor_topic, Y_lr)
  pair_keep  <- anchor_obj$pair_keep
  pair_names <- anchor_obj$pair_names_anchor
  pair_info  <- anchor_obj$pair_info

  coef_sub   <- coef_mat[pair_keep, tp_coefs, drop = FALSE]
  signif_sub <- signif_mat[pair_keep, tp_coefs, drop = FALSE]

  rownames(coef_sub)   <- pair_names
  rownames(signif_sub) <- pair_names

  df_out <- expand.grid(
    pair = pair_names,
    coef_name = tp_coefs,
    stringsAsFactors = FALSE
  )

  df_out <- merge(df_out, pair_info, by = "pair", all.x = TRUE)

  df_out$estimate_raw <- mapply(function(p, c) coef_sub[p, c], df_out$pair, df_out$coef_name)
  df_out$estimate <- ifelse(df_out$flip_sign, -df_out$estimate_raw, df_out$estimate_raw)
  df_out$signif <- mapply(function(p, c) signif_sub[p, c], df_out$pair, df_out$coef_name)
  df_out$time <- unname(time_map[df_out$coef_name])

  df_out
}

plot_anchor_df <- function(df_in, anchor_topic, method_label, signif_label) {
  ggplot(df_in, aes(x = time, y = estimate, group = pair_label)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
    geom_line(color = "black") +
    geom_point(aes(fill = signif), shape = 21, color = "black", size = 1.5) +
    facet_wrap(~ pair_label, scales = "free_y", ncol = 4) +
    scale_fill_manual(
      values = c("TRUE" = "black", "FALSE" = "white"),
      name = signif_label
    ) +
    labs(
      title = paste0("Anchor topic T", anchor_topic, " vs all others (", method_label, ")"),
      x = "Time",
      y = paste0("Estimated effect on log(T", anchor_topic, "/other) vs d0")
    ) +
    theme_bw(base_size = 13)
}
```

## 7. Anchor topic plots for T1 and T7
```{r}
common_tp_anchor <- Reduce(intersect, list(
  tp_coefs,
  colnames(fit_eb$coefficients),
  colnames(coef_hat),
  colnames(mr_coef_mat),
  colnames(sig_limma),
  colnames(sig_ash),
  colnames(sig_epb),
  colnames(sig_mr)
))

for (anchor_topic in c(1, 7)) {
  df_limma <- make_anchor_df(anchor_topic, Y_lr, fit_eb$coefficients, sig_limma, common_tp_anchor, time_map)
  print(plot_anchor_df(df_limma, anchor_topic, "limma", "BH < 0.05"))

  df_ash <- make_anchor_df(anchor_topic, Y_lr, coef_hat, sig_ash, common_tp_anchor, time_map)
  print(plot_anchor_df(df_ash, anchor_topic, "ASH", "ASH q < 0.05"))

  df_epb <- make_anchor_df(anchor_topic, Y_lr, coef_hat, sig_epb, common_tp_anchor, time_map)
  print(plot_anchor_df(df_epb, anchor_topic, "EPB", "EPB adj. p < 0.05"))

  df_mr <- make_anchor_df(anchor_topic, Y_lr, mr_coef_mat, sig_mr, common_tp_anchor, time_map)
  print(plot_anchor_df(df_mr, anchor_topic, "MR.ASHR", "lfsr < 0.05"))
}
```

