---
title: "RNA3.2"
output: pdf_document
date: "2026-03-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(limma)
library(dplyr)
library(tidyr)
library(matrixStats)
library(ggplot2)
library(stringr)
```

## pairwise log comparision: treatment*dosage*time*cell_state
### 1. preprocessing
```{r}
alpha <- 0.05
topics_per_plot <- 4

# metadata
df <- droplevels(meta_balanced) %>%
  mutate(
    treatment  = droplevels(factor(treatment)),
    dosage     = droplevels(factor(dosage)),
    time       = droplevels(factor(time)),
    cell_state = droplevels(factor(cell_state, levels = c("G1","S","G2","M","APOP")))
  )

# topic proportions
theta <- fit_multinom$L
stopifnot(nrow(theta) == nrow(df))

N <- nrow(theta)
K <- ncol(theta)

# topic-specific epsilon
c0 <- 1e-4
eps_k <- c0 * colMedians(theta, na.rm = TRUE)
eps_k <- pmin(pmax(eps_k, 1e-12), 1e-2)

theta01   <- pmin(pmax(theta, 0), 1)
log_theta <- log(sweep(theta01, 2, eps_k, FUN = "+"))

# combined condition
df$cond_cs <- interaction(
  df$treatment, df$dosage, df$time, df$cell_state,
  sep = "_", drop = TRUE
)
df$cond_cs <- relevel(df$cond_cs, ref = "WT_0_0_G1")

design_cs <- model.matrix(~ cond_cs, data = df)
colnames(design_cs) <- make.names(colnames(design_cs))
stopifnot(limma::is.fullrank(design_cs))
```


---

### 2. Build pairwise features and fit once


```{r}
parse_pairs <- function(pair_vec) {
  parts <- do.call(rbind, strsplit(pair_vec, "_vs_"))
  tibble(
    pair = pair_vec,
    k1 = as.integer(sub("^T", "", parts[, 1])),
    k2 = as.integer(sub("^T", "", parts[, 2]))
  )
}

build_pairwise_results <- function(log_theta, design, robust = TRUE) {
  K <- ncol(log_theta)
  pair_mat <- t(combn(K, 2))
  n_pairs  <- nrow(pair_mat)

  log_ratio_mat <- matrix(NA_real_, nrow = nrow(log_theta), ncol = n_pairs)
  for (p in seq_len(n_pairs)) {
    k1 <- pair_mat[p, 1]
    k2 <- pair_mat[p, 2]
    log_ratio_mat[, p] <- log_theta[, k1] - log_theta[, k2]
  }

  colnames(log_ratio_mat) <- apply(pair_mat, 1, function(idx) {
    paste0("T", idx[1], "_vs_T", idx[2])
  })

  Y_lr <- t(log_ratio_mat)

  fit_lr <- lmFit(Y_lr, design)
  fit_lr <- eBayes(fit_lr, robust = robust)

  coef_names <- colnames(fit_lr$coefficients)
  cond_cs_coefs <- grep("^cond_cs", coef_names, value = TRUE)

  res_pairs_all <- bind_rows(lapply(cond_cs_coefs, function(cf) {
    tt <- topTable(fit_lr, coef = cf, number = Inf, sort.by = "none")
    tt$coef <- cf
    tt$pair <- rownames(tt)
    tt
  }))

  pair_info <- parse_pairs(rownames(Y_lr))

  res_long <- res_pairs_all %>%
    left_join(pair_info, by = "pair") %>%
    select(coef, pair, logFC, adj.P.Val, P.Value, k1, k2)

  res_directed <- bind_rows(
    res_long %>% transmute(coef, pair, topic = k1, other = k2,
                           beta =  logFC, q = adj.P.Val, p = P.Value),
    res_long %>% transmute(coef, pair, topic = k2, other = k1,
                           beta = -logFC, q = adj.P.Val, p = P.Value)
  )

  list(
    Y_lr = Y_lr,
    fit_lr = fit_lr,
    cond_cs_coefs = cond_cs_coefs,
    res_pairs_all = res_pairs_all,
    res_long = res_long,
    res_directed = res_directed
  )
}

pw <- build_pairwise_results(log_theta, design_cs, robust = TRUE)
res_long <- pw$res_long
res_directed <- pw$res_directed
```


---

### 3. Helpers for coefficient parsing


```{r}
parse_coef_fields <- function(dat) {
  dat %>%
    mutate(coef_raw = sub("^cond_cs", "", coef)) %>%
    separate(
      coef_raw,
      into = c("treatment", "dosage", "time", "cell_state"),
      sep = "_",
      remove = FALSE,
      fill = "right"
    ) %>%
    mutate(
      dosage = factor(dosage, levels = c("10", "20", "30")),
      time = as.numeric(time),
      cell_state = factor(cell_state, levels = c("G1","S","G2","M","APOP"))
    )
}
```


---

### 4. Least-extreme summary


```{r}
least_extreme <- res_directed %>%
  group_by(coef, topic) %>%
  summarise(
    S = min(abs(beta), na.rm = TRUE),
    Q_worst = max(q, na.rm = TRUE),
    frac_sig = mean(q < alpha, na.rm = TRUE),
    other_min = other[which.min(abs(beta))][1],
    beta_min = beta[which.min(abs(beta))][1],
    .groups = "drop"
  ) %>%
  mutate(
    topic = paste0("T", topic),
    other_min = paste0("T", other_min)
  )

plot_le <- least_extreme %>%
  parse_coef_fields() %>%
  filter(treatment != "WT") %>%
  mutate(topic_id = as.integer(sub("^T", "", topic)))
  
```


#### Line plots for least-extreme effect size


```{r}
make_topic_block_plot <- function(dat, yvar, ylab, ttl_prefix, lo, hi) {
  ggplot(
    dat %>% filter(topic_id >= lo, topic_id <= hi),
    aes(x = time, y = .data[[yvar]], color = treatment, linetype = dosage)
  ) +
    geom_line() +
    facet_grid(cell_state ~ topic, scales = "free_y") +
    scale_x_log10() +
    theme_bw() +
    labs(
      x = "Time",
      y = ylab,
      title = paste0(ttl_prefix, " (T", lo, "–T", hi, ")")
    )
}

p1 <- make_topic_block_plot(
  plot_le, "S",
  "Least-extreme |beta| = min over other topics",
  "Topic distinctness over time", 1, 8
)

p2 <- make_topic_block_plot(
  plot_le, "S",
  "Least-extreme |beta| = min over other topics",
  "Topic distinctness over time", 9, 16
)

p1
p2

```



#### Heatmaps for fraction significant and least-extreme effect
```{r}
topic_summary <- res_directed %>%
  group_by(coef, topic) %>%
  summarise(
    n_total = sum(!is.na(q)),
    n_sig = sum(q < alpha, na.rm = TRUE),
    frac_sig = n_sig / n_total,
    S = min(abs(beta), na.rm = TRUE),
    other_min = other[which.min(abs(beta))][1],
    beta_min = beta[which.min(abs(beta))][1],
    Q_worst = max(q, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    topic = paste0("T", topic),
    other_min = paste0("T", other_min)
  )

plot_df2 <- topic_summary %>%
  parse_coef_fields() %>%
  mutate(
    time = factor(time, levels = sort(unique(time)))
  ) %>%
  filter(treatment != "WT")

base_topic_heatmap <- function(dat, fill_var, fill_lab, ttl) {
  ggplot(dat, aes(x = time, y = topic, fill = .data[[fill_var]])) +
    geom_tile() +
    facet_grid(cell_state ~ treatment + dosage, scales = "free_x") +
    theme_bw() +
    theme(
      axis.text.x  = element_text(size = 8, angle = 45, hjust = 1),
      axis.title.x = element_text(size = 9),
      axis.title.y = element_text(size = 9),
      strip.text.y = element_text(size = 8),
      strip.text.x = element_text(size = 7)
    ) +
    labs(x = "Time", y = "Topic", fill = fill_lab, title = ttl)
}

p_G1S_frac <- base_topic_heatmap(
  filter(plot_df2, cell_state %in% c("G1","S")),
  "frac_sig", "Fraction\nsignificant", "Fraction significant (G1, S)"
)

p_G2M_frac <- base_topic_heatmap(
  filter(plot_df2, cell_state %in% c("G2","M")),
  "frac_sig", "Fraction\nsignificant", "Fraction significant (G2, M)"
)

p_APOP_frac <- base_topic_heatmap(
  filter(plot_df2, cell_state == "APOP"),
  "frac_sig", "Fraction\nsignificant", "Fraction significant (APOP)"
)

p_G1S_S <- base_topic_heatmap(
  filter(plot_df2, cell_state %in% c("G1","S")),
  "S", "Least-extreme\n|beta|", "Topic distinctness (G1, S)"
)

p_G2M_S <- base_topic_heatmap(
  filter(plot_df2, cell_state %in% c("G2","M")),
  "S", "Least-extreme\n|beta|", "Topic distinctness (G2, M)"
)

p_APOP_S <- base_topic_heatmap(
  filter(plot_df2, cell_state == "APOP"),
  "S", "Least-extreme\n|beta|", "Topic distinctness (APOP)"
)

p_G1S_frac
p_G2M_frac
p_APOP_frac

p_G1S_S
p_G2M_S
p_APOP_S
```


---

### 5. Pairwise beta plot for one topic

```{r}
topic_k <- 15

plot_topic_beta <- res_directed %>%
  filter(topic == topic_k, other != topic_k) %>%
  parse_coef_fields() %>%
  mutate(
    other_id = other,
    other = paste0("T", other),
    other = factor(other, levels = paste0("T", setdiff(seq_len(K), topic_k))),
    topic = paste0("T", topic_k),
    sig = q < alpha
  ) %>%
  filter(treatment != "WT")

make_beta_plot <- function(dat, ttl) {
  ggplot(
    dat,
    aes(x = time, y = beta,
        group = interaction(treatment, dosage),
        color = treatment, linetype = dosage)
  ) +
    geom_hline(yintercept = 0, linewidth = 0.3) +
    geom_line(linewidth = 0.7) +
    geom_point(aes(shape = sig), size = 1.8) +
    scale_shape_manual(values = c(`FALSE` = 1, `TRUE` = 16)) +
    facet_grid(cell_state ~ other, scales = "free_y") +
    scale_x_log10(breaks = c(5, 10, 30, 240, 1440)) +
    theme_bw() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
      strip.text.x = element_text(size = 7),
      strip.text.y = element_text(size = 8)
    ) +
    labs(
      x = "Time",
      y = expression(beta[c]),
      shape = "BH < 0.05",
      title = ttl
    )
}

p_1_8 <- make_beta_plot(
  filter(plot_topic_beta, other_id %in% 1:8),
  "T15 vs other topics (T1–T8)"
)

p_9_16 <- make_beta_plot(
  filter(plot_topic_beta, other_id %in% 9:16),
  "T15 vs other topics (T9–T16)"
)

p_1_8
p_9_16
```


---

### 6. Most-extreme summary


```{r}
most_extreme <- res_directed %>%
  group_by(coef, topic) %>%
  summarise(
    S_max = max(abs(beta), na.rm = TRUE),
    Q_best = min(q, na.rm = TRUE),
    frac_sig = mean(q < alpha, na.rm = TRUE),
    other_max = other[which.max(abs(beta))][1],
    beta_max = beta[which.max(abs(beta))][1],
    .groups = "drop"
  ) %>%
  mutate(
    topic = paste0("T", topic),
    other_max = paste0("T", other_max)
  )

plot_me <- most_extreme %>%
  parse_coef_fields() %>%
  filter(treatment != "WT") %>%
  mutate(topic_id = as.integer(sub("^T", "", topic)))
```


#### Line plots for most-extreme effect size

```{r}
p1_max <- make_topic_block_plot(
  plot_me, "S_max",
  "Most-extreme |beta| = max over other topics",
  "Topic extreme-contrast over time", 1, 8
)

p2_max <- make_topic_block_plot(
  plot_me, "S_max",
  "Most-extreme |beta| = max over other topics",
  "Topic extreme-contrast over time", 9, 16
)

p1_max
p2_max
```


#### Heatmaps for signed and absolute most-extreme effects

```{r}
plot_df_max <- most_extreme %>%
  parse_coef_fields() %>%
  mutate(
    time = factor(time, levels = sort(unique(time)))
  ) %>%
  filter(treatment != "WT")

base_beta_max_heatmap <- function(dat, fill_var, fill_lab, ttl, signed = FALSE) {
  p <- ggplot(dat, aes(x = time, y = topic, fill = .data[[fill_var]])) +
    geom_tile() +
    facet_grid(cell_state ~ treatment + dosage, scales = "free_x") +
    theme_bw() +
    theme(
      axis.text.x  = element_text(size = 8, angle = 45, hjust = 1),
      axis.title.x = element_text(size = 9),
      axis.title.y = element_text(size = 9),
      strip.text.y = element_text(size = 8),
      strip.text.x = element_text(size = 7)
    ) +
    labs(x = "Time", y = "Topic", fill = fill_lab, title = ttl)

  if (signed) {
    p <- p + scale_fill_gradient2(name = fill_lab, midpoint = 0)
  }
  p
}

p_G1S_beta <- base_beta_max_heatmap(
  filter(plot_df_max, cell_state %in% c("G1","S")),
  "beta_max", "beta_max\n(signed)", "Most-extreme signed effect (G1, S)", TRUE
)

p_G2M_beta <- base_beta_max_heatmap(
  filter(plot_df_max, cell_state %in% c("G2","M")),
  "beta_max", "beta_max\n(signed)", "Most-extreme signed effect (G2, M)", TRUE
)

p_APOP_beta <- base_beta_max_heatmap(
  filter(plot_df_max, cell_state == "APOP"),
  "beta_max", "beta_max\n(signed)", "Most-extreme signed effect (APOP)", TRUE
)

p_G1S_mag <- base_beta_max_heatmap(
  filter(plot_df_max, cell_state %in% c("G1","S")),
  "S_max", "Most extreme\n|beta|", "Topic extreme-contrast magnitude (G1, S)"
)

p_G2M_mag <- base_beta_max_heatmap(
  filter(plot_df_max, cell_state %in% c("G2","M")),
  "S_max", "Most extreme\n|beta|", "Topic extreme-contrast magnitude (G2, M)"
)

p_APOP_mag <- base_beta_max_heatmap(
  filter(plot_df_max, cell_state == "APOP"),
  "S_max", "Most extreme\n|beta|", "Topic extreme-contrast magnitude (APOP)"
)

p_G1S_beta
p_G2M_beta
p_APOP_beta

p_G1S_mag
p_G2M_mag
p_APOP_mag
```
