---
title: "Method B ILR (standard basis) + linear models"
output: pdf_document
date: "2025-10-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# ============================================================
# Synthetic simulation: condition-dependent topic model
# ============================================================
# Description:
#   Generates single-cell-like counts Y_{ng} under a
#   Gamma–Poisson (Negative Binomial) model with topic
#   proportions depending on condition (e.g., drug vs no drug).
# ============================================================

simulate_topics <- function(
  N = 1000,        # number of cells
  G = 500,         # number of genes
  K = 5,           # number of topics
  frac_diff = 0.4, # fraction of topics with nonzero beta
  beta_strength = 2, # effect size for condition-dependent topics
  seed = 123
) {
  set.seed(seed)

  # -------------------------
  # 1️⃣ Experimental condition
  # -------------------------
  X <- rbinom(N, 1, 0.5)  # 0 = control, 1 = drug

  # -------------------------
  # 2️⃣ Topic-level parameters
  # -------------------------
  mu_k   <- rnorm(K, 0, 1)
  diff_k <- sample(c(TRUE, FALSE), K, replace = TRUE,
                   prob = c(frac_diff, 1 - frac_diff))
  beta_k <- ifelse(diff_k, rnorm(K, beta_strength, 0.5), 0)
  diff_topic_ids <- which(diff_k)


  # -------------------------
  # 3️⃣ Latent topic logits η_{nk}
  # -------------------------
  eta <- matrix(NA, N, K)
  for (k in 1:K) {
    eta[, k] <- rnorm(N, mean = mu_k[k] + X * beta_k[k], sd = 1)
  }

  # -------------------------
  # 4️⃣ Softmax to get topic proportions z_{nk}
  # -------------------------
  softmax <- function(v) exp(v) / sum(exp(v))
  Z <- t(apply(eta, 1, softmax))  # N x K
  stopifnot(all(abs(rowSums(Z) - 1) < 1e-8))

  # -------------------------
  # 5️⃣ Topic–gene weights w_{kg}
  # -------------------------
  T_mat <- matrix(rlnorm(K * G, meanlog = 0, sdlog = 1), K, G)
  W <- t(apply(T_mat, 1, function(v) v / sum(v)))  # normalize rows to sum 1

  # -------------------------
  # 6️⃣ Library size per cell l_n
  # -------------------------
  l_n <- runif(N, 1000, 10000)

  # -------------------------
  # 7️⃣ Gene-specific dispersion α_g
  # -------------------------
  alpha_g <- rexp(G, rate = 0.2)  # mean = 5

  # -------------------------
  # 8️⃣ Latent mean expression x_{ng}
  # -------------------------
  mean_ng <- Z %*% W  # expected fraction per gene (N x G)
  mean_ng <- mean_ng * l_n  # scale by library size

  # Gamma–Poisson (Negative Binomial)
  X_ng <- matrix(NA, N, G)
  for (g in 1:G) {
    shape <- alpha_g[g]
    rate  <- alpha_g[g] / mean_ng[, g]
    X_ng[, g] <- rgamma(N, shape = shape, rate = rate)
  }

  # -------------------------
  # 9️⃣ Observed counts Y_{ng}
  # -------------------------
  Y_ng <- matrix(rpois(N * G, lambda = as.vector(X_ng)), N, G)

  # -------------------------
  # 🔟 Return
  # -------------------------
  list(
    Y = Y_ng,          # counts (N x G)
    Z = Z,             # topic proportions (N x K)
    W = W,             # topic–gene weights (K x G)
    beta_true = beta_k,          # true beta per topic
    diff_topics = diff_k,        # logical: which topics differ
    diff_topic_ids = diff_topic_ids,  # numeric indices of diff topics
    X = X,              # condition (0/1)
    mu = mu_k,
    alpha_g = alpha_g,
    library_size = l_n
  )

  # list(
  #   Y = Y_ng,          # counts (N x G)
  #   Z = Z,             # topic proportions (N x K)
  #   W = W,             # topic–gene weights (K x G)
  #   beta_true = beta_k,
  #   diff_topics = diff_k,
  #   X = X,
  #   mu = mu_k,
  #   alpha_g = alpha_g,
  #   library_size = l_n
  # )
}

# ============================================================
# Example usage
# ============================================================
sim <- simulate_topics(N = 800, G = 300, K = 6, frac_diff = 0.5, beta_strength = 2)
sim$diff_topic_ids
str(sim)
summary(rowSums(sim$Y))
table(sim$diff_topics)

# Example: visualize first few cells’ topic proportions by condition
boxplot(sim$Z[, 1] ~ sim$X, main = "Topic 1 proportion by condition",
        names = c("No Drug", "Drug"), ylab = "z_{n1}")

```
Y — an N×G matrix of observed counts.

Z — an N×K matrix of topic proportions for each cell, which depend on the experimental condition.

W — a K×G matrix of topic–gene weights describing how genes load onto each topic.

beta_true — the true β values for each topic that control how strongly the topic responds to the condition.

diff_topics — a logical vector indicating which topics are condition-dependent (β ≠ 0).

X — the binary condition variable (0 = control, 1 = drug).

library_size — a vector giving the sequencing depth for each cell.

alpha_g — the gene-level dispersion parameter controlling overdispersion in the Gamma–Poisson model.

```{r}
diff_topics <- sample(K, size = round(frac_diff * K))
beta_true <- rep(0, K)
beta_true[diff_topics] <- rnorm(length(diff_topics), mean = beta_strength, sd = 0.2)

```

## Method B
```{r}
# Core compositional + modeling packages
library(compositions)  # for ilr()
library(limma)         # for moderated linear models
library(dplyr)

Z_true <- sim$Z        # topic proportions
X_cond <- sim$X        # condition (0/1)
K <- ncol(Z_true)
# ILR
Z_ilr <- ilr(acomp(Z_true))   # (N × (K-1)) ILR coordinates
colnames(Z_ilr) <- paste0("ILR", seq_len(K - 1))
dim(Z_ilr)
# 
design <- model.matrix(~ X_cond)
colnames(design) <- c("Intercept", "ConditionDrug")
# Fit linear model
fit <- lmFit(t(Z_ilr), design)   # topics/coordinates in rows, samples in columns
fit <- eBayes(fit)
res <- topTable(fit, coef = "ConditionDrug", number = Inf)
head(res)


```

```{r}
sig <- subset(res, adj.P.Val < 0.05)
sig

```
# Z_ILR =log(Z)⋅V, V is basis
```{r}
basis <- ilrBase(acomp(Z_true))
rownames(basis) <- paste0("Topic", seq_len(K))
colnames(basis) <- paste0("ILR", seq_len(K - 1))
basis

```

```{r}
# ============================================================
# Method B — ILR (standard basis) + linear models
# ============================================================

library(compositions)  # ilr(), acomp(), ilrBase()
library(limma)
library(dplyr)
library(ggplot2)

# --- 1️⃣ Simulate data --------------------------------------
sim <- simulate_topics(N = 800, G = 300, K = 6, frac_diff = 0.4, beta_strength = 2)
Z_true <- sim$Z     # topic proportions
X_cond <- sim$X     # condition (0 = control, 1 = drug)
K <- ncol(Z_true)

# --- 2️⃣ ILR transform --------------------------------------
Z_ilr <- ilr(acomp(Z_true))
colnames(Z_ilr) <- paste0("ILR", seq_len(K - 1))

# --- 3️⃣ Design matrix --------------------------------------
design <- model.matrix(~ X_cond)
colnames(design) <- c("Intercept", "ConditionDrug")

# --- 4️⃣ Fit moderated linear models ------------------------
fit <- lmFit(t(Z_ilr), design)
fit <- eBayes(fit)
res <- topTable(fit, coef = "ConditionDrug", number = Inf)
print(res)

# --- 5️⃣ Identify significant ILR coordinates ---------------
sig <- subset(res, adj.P.Val < 0.05)
sig

# --- 6️⃣ Interpret back in topic space -----------------------
basis <- ilrBase(acomp(Z_true))
rownames(basis) <- paste0("Topic", seq_len(K))
colnames(basis) <- paste0("ILR", seq_len(K - 1))
basis

# Interpretation guide:
# Each column = ILR coordinate (balance)
# Positive weights  → topics in numerator of that balance
# Negative weights → topics in denominator

```
```{r}
plot_df <- data.frame(
  Condition = factor(X_cond, labels = c("No Drug", "Drug")),
  ILR1 = Z_ilr[, 1]
)

ggplot(plot_df, aes(x = Condition, y = ILR1, fill = Condition)) +
  geom_boxplot(alpha = 0.6) +
  labs(title = "ILR₁ coordinate by condition",
       y = "ILR₁ (log-ratio balance)") +
  theme_classic()

```
```{r}
library(vegan)

# PERMANOVA using Euclidean distances in ILR space
perm_global <- adonis2(Z_ilr ~ X_cond, method = "euclidean")
perm_global

```

```{r}
manova_res <- manova(Z_ilr ~ X_cond)
summary(manova_res, test = "Wilks")

```

```{r}
# Number of significant ILR coordinates at FDR < 0.05
n_disc <- sum(res$adj.P.Val < 0.05)
n_disc

# Which coordinates were discovered
sig_coords <- rownames(res)[res$adj.P.Val < 0.05]
sig_coords


```
```{r}
library(compositions)
library(limma)

run_methodB <- function(Z, X, fdr = 0.05) {
  K <- ncol(Z)
  Z_ilr <- ilr(acomp(Z))
  design <- model.matrix(~ X)                 # Intercept + condition
  fit <- lmFit(t(Z_ilr), design) |> eBayes()
  tt  <- topTable(fit, coef = 2, number = Inf)  # coef=2 is the condition term

  # per-coordinate discoveries at FDR q
  disc_idx <- which(tt$adj.P.Val < fdr)
  disc_names <- rownames(tt)[disc_idx]

  # ILR basis for interpretation (topics × ILR coords)
  basis <- ilrBase(acomp(Z))
  rownames(basis) <- paste0("Topic", seq_len(K))
  colnames(basis) <- paste0("ILR", seq_len(K-1))

  list(
    Z_ilr = Z_ilr,
    design = design,
    fit = fit,
    table = tt,
    n_disc = length(disc_idx),
    disc_names = disc_names,
    basis = basis
  )
}

```

```{r}
out <- run_methodB(sim$Z, sim$X, fdr = 0.05)
out$n_disc
out$disc_names
out$basis

```
```{r}
global_tests <- function(Z_ilr, X) {
  res <- list()
  if (!requireNamespace("vegan", quietly = TRUE))
    stop("Install 'vegan' for PERMANOVA: install.packages('vegan')")
  res$permanova <- vegan::adonis2(Z_ilr ~ X, method = "euclidean")
  res$manova    <- summary(manova(Z_ilr ~ X), test = "Wilks")
  res
}
g <- global_tests(out$Z_ilr, sim$X)
g$permanova      # global p-value
g$manova

```
## 3) Stability under resampling (bootstrap cells)
```{r}
stability_methodB <- function(Z, X, fdr = 0.05, B = 50, seed = 1) {
  set.seed(seed)
  K <- ncol(Z); p <- K - 1
  stab <- numeric(p)

  for (b in 1:B) {
    idx <- sample(seq_len(nrow(Z)), replace = TRUE)
    ob  <- run_methodB(Z[idx, , drop=FALSE], X[idx], fdr = fdr)
    sig <- rep(0L, p); names(sig) <- rownames(ob$table)
    sig[ob$disc_names] <- 1L
    stab <- stab + sig
  }
  stab / B  # fraction of bootstraps where each ILR coordinate is significant
}

stab <- stability_methodB(sim$Z, sim$X, fdr = 0.05, B = 50)
# # names (ILR indices)
names(stab) <- paste0("ILR", seq_along(stab))
stab


```



```{r}
sim_null <- simulate_topics(N = 800, G = 300, K = 6, frac_diff = 0.0, beta_strength = 0)
Z_ilr_null <- ilr(acomp(sim_null$Z))

design_null <- model.matrix(~ sim_null$X)
fit_null <- lmFit(t(Z_ilr_null), design_null) |> eBayes()
res_null <- topTable(fit_null, coef = "sim_null$X1", number = Inf)

hist(res_null$P.Value, breaks = 30,
     main = "Type I calibration under null",
     xlab = "p-value")
mean(res_null$adj.P.Val < 0.05)   # empirical false-positive rate

```

## 4) Type I error calibration (null sims) and Power
```{r}
# assumes your simulate_topics() is in scope
# If there’s no difference between conditions, how often does Method B report a significant ILR coordinate
type1_calibration <- function(R = 200, N = 800, G = 300, K = 6, fdr = 0.05) {
  fp <- numeric(R)
  for (r in 1:R) {
    sim0 <- simulate_topics(N=N, G=G, K=K, frac_diff=0, beta_strength=0,seed = r)
    tt0  <- run_methodB(sim0$Z, sim0$X, fdr=fdr)$table
    fp[r] <- mean(tt0$adj.P.Val < fdr)
  }
  list(mean_fp = mean(fp), fp = fp)
}
# When there is a true signal, how often does Method B detect it?
power_estimate <- function(R = 200, N = 800, G = 300, K = 6,
                           frac_diff = 0.5, beta_strength = 2, fdr = 0.05) {
  det_any <- logical(R)
  for (r in 1:R) {
    sim1 <- simulate_topics(N=N, G=G, K=K,
                            frac_diff=frac_diff, beta_strength=beta_strength,seed = r)

    out1 <- run_methodB(sim1$Z, sim1$X, fdr=fdr)
    det_any[r] <- out1$n_disc > 0
  }
  list(power = mean(det_any), detect_any = det_any)
}

# Type I calibration (should be near 0.05 if well-calibrated)
t1 <- type1_calibration(R = 100)
t1$mean_fp

# Power at a given effect size
pw <- power_estimate(R = 2, frac_diff = 0.5, beta_strength = 2)
pw$power

```

5) Quick p-value diagnostic plot (null)
```{r}
plot_pvals_null <- function(R = 200, N = 800, G = 300, K = 6) {
  sim0 <- simulate_topics(N=N, G=G, K=K, frac_diff=0, beta_strength=0)
  tt0  <- run_methodB(sim0$Z, sim0$X)$table
  hist(tt0$P.Value, breaks = 30, main = "Null p-values (Method B)",
       xlab = "p-value")
}
plot_pvals_null()

```
```{r}
out <- run_methodB(sim$Z, sim$X, fdr = 0.05)
g   <- global_tests(out$Z_ilr, sim$X)
stab <- stability_methodB(sim$Z, sim$X, B = 50)

out$n_disc
g$permanova
stab

```


```{r}
n_sim <- 100
detect_any <- logical(n_sim)

for (i in seq_len(n_sim)) {
  sim_i <-simulate_topics(N = 800, G = 300, K = 6, frac_diff = 0.5, beta_strength = 2)
  Z_ilr_i <- ilr(acomp(sim_i$Z))
  design_i <- model.matrix(~ sim_i$X)
  fit_i <- lmFit(t(Z_ilr_i), design_i) |> eBayes()
  res_i <- topTable(fit_i, coef = 2, number = Inf)
  detect_any[i] <- any(res_i$adj.P.Val < 0.05)
}

mean(detect_any)  # estimated power

```

```{r}
set.seed(1)
boot_ids <- replicate(50, sample(seq_len(nrow(Z_ilr)), replace = TRUE), simplify = FALSE)

stable_counts <- sapply(boot_ids, function(idx) {
  fit_b <- lmFit(t(Z_ilr[idx, ]), design[idx, , drop = FALSE]) |> eBayes()
  res_b <- topTable(fit_b, coef = 2, number = Inf)
  as.integer(res_b$adj.P.Val < 0.05)
})

# Stability: fraction of bootstraps where each coordinate is significant
stability <- rowMeans(stable_counts)
stability

```

